A semantics of core AUTOSAR

### 2016-05-15 paper/paper.tex
Rename U to P + small fixes

%TODO: port name convention change from U to P? (see ../prolog/semantics.eprover)
%
\[
\begin{array}{rcll}
	I	& \in		& \text{Component names} \\
	R	& \in		& \text{Runnable names} \\
	U	& \in		& \text{Port names} \\
	E	& \in		& \text{Sender-receiver element names} \\
	O	& \in		& \text{Client-server operation names} \\
	S	& \in		& \text{Inter-runnable variable names} \\
	X	& \in		& \text{Exclusive area names} \\
	\\
	A	& \in		& \text{Addresses} \\
	A	& ::=		& \Tr{R}{I}   \sep   \Tep{E}{P}{I}   \sep \Top{O}{P}{I}   \sep   \Ts{S}{I}   \sep   \Tx{X}{I}   \\
\end{array}
\]
%
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{TODO2}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\runnable{\Tr{R}{I}}{TODOK}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\rinst{\Tr{R}{I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\excl{\Tx{X}{I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\irv{\Ts{S}{I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\delem{\Tep{E}{U}{I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\qelem{\Tep{E}{U}{I}}{\Vs}{TODO}$ \newline
Queued sender-receiver data element $E$ in port $U$ of component $I$, currently holding value sequence $\Vs$.
\item $\opres{\Top{O}{U}{I}}{\Vs}$ \newline
Client-server operation $O$ in port $U$ of component $I$, currently holding result value sequence $\Vs$.
\item $\timer{\Tr{R}{I}}{T}$ \newline  %TODO: timer has only 2 args in the semantics so {T_p} is not described
Countdown timer state for runnable $R$ of component $I$ with $T$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
$$
\begin{array}{rcll}
	V	& \in		& \text{Values} \\
	N	& \in		& \text{Natural numbers} \\
	B	& \in		& \text{Boolean values} \\
	T	& \in		& \text{Time values} \\
	C	& ::=		& \To{O}{U}{I} \sep \Tvoid \\
	\Act	& ::=		& \Tidle \sep \Tpending \sep \serving{\Cs}{\Vs} \\
\end{array}
$$
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

The code part of a runnable instance should technically be a representation of the C or Matlab function that must accompany an AUTOSAR runnable definition once it is complete. However, since our task in this report is not to investigate the semantics of these languages in every detail, a more abstract notion of executable code will be required. To this end, we have chosen to ignore all details of internal computations, and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. Because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style to allow any data-dependencies to be expressed.
%TODO: several LaTeX commands here are used with more arguments than in the semantics below
$$
\begin{array}{lll}
	\Code	& ::=	& \Enter{X}{\Cont}			\\
			& |	& \Exit{X}{\Cont}			\\
			& |	& \IrvWrite{S}{V}{\Cont}		\\
			& |	& \IrvRead{S}{\Cont}			\\
			& |	& \Send{E}{U}{V}{\Cont}		\\
			& |	& \Receive{E}{U}{\Cont}		\\
			& |	& \Write{E}{U}{V}{\Cont}		\\
			& |	& \Read{E}{U}{\Cont}		\\
			& |	& \Invalidate{E}{U}{\Cont}		\\
			& |	& \IsUpdated{E}{U}{\Cont}		\\
			& |	& \Call{O}{U}{V}{\Cont}		\\
			& |	& \Result{O}{U}{\Cont}		\\
			& |	& \Return{V}				\\
	\\
	\Cont	& \in	& V \rightarrow \Code		\\
\end{array}
$$

Together with an operator for composing process terms in parallel, the process grammar for our AUTOSAR calculus becomes
$$
\begin{array}{rcll}
	P,Q 	& ::= &
		\runnable{A}{K}{T}{\Act}{N}	\\
		& | &
		\rinst{A}{C}{\Xs}{\Code}	\\
		& | &
		\excl{A}{B}			\\
		& | &
		\irv{A}{V}				\\
		& | &
		\delem{A}{B}{V}			\\
		& | &
		\qelem{A}{\Vs}{TODO}			\\
		& | &
		\opres{A}{\Vs}			\\
		& | &
		\timer{A}{T}		\\
		& | &
		P|Q					\\
\end{array}
$$

TODO: links from syntax and concepts introduced here to the AUTSAR spec

@@

\item For each runnable $R$ of each component $I$, a term
$$
\begin{array}{c}
	\runnable{\Tr{R}{I}}{TODO}{0.0}{\Act}{0}
\end{array}
$$
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\Tpending$ if $R$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\Tidle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area $X$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\excl{\Tx{X}{I}}{\Tfalse}
\end{array}
$$

\item For each inter-runnable variable $S$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\irv{\Ts{S}{I}}{V}
\end{array}
$$
where $V$ is the initial value attributed to $S$ if it exists, otherwise $V$ is $\Tvoid$.

\item For each data element $E$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\qelem{\Tep{E}{U}{I}}{\epsilon}{TODO}
\end{array}
$$
if the \emph{swImplPolicy} attribute of element $E$ is \emph{queued}; otherwise, a term
$$
\begin{array}{c}
	\delem{\Tep{E}{U}{I}}{\Tfalse}{\text{nodata}}
\end{array}
$$

\item For each operation $O$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\opres{\Tep{O}{U}{I}}{\epsilon}
\end{array}
$$

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\timer{\Tr{R}{I}}{0.0}
\end{array}
$$
% where $T$ is the period of the event.
%TODO: Does this mean all timed runnables will want to act _immediately_ (at the same time)? No init delay?

@@

Some further static model information will also be referenced by the transition rules. We refer to the \emph{connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
$$
	\Tconnect{\Tep{E}{$U_1$}{$I_1$}}{\Tep{E}{$U_2$}{$I_2$}} \;\;\text{for all elements $E$ of port $U_2$}
$$
if there is a connector in the model from provided sender-receiver port $U_1$ of component $I_1$ to required port $U_2$ of component $I_2$; and
$$
	\Tconnect{\Top{O}{$U_1$}{$I_1$}}{\Top{O}{$U_2$}{$I_2$}} \;\;\text{for all operations $O$ of port $U_1$}
$$
if there is a connector in the model from required client-server port $U_1$ of component $I_1$ to provided port $U_2$ of component $I_2$.

@@

%TODO: Patrik got here 2016-04-27 in a reading + macro conversion round
\subsection{Exclusive areas}

@@

%\[
%\begin{array}{r@{}c@{}llllll}

%\input{../prolog/tool/semantics.tex}

>>>>

%
\[
\begin{array}{rcll}
  I & \in & \text{Component names} \\
  R & \in & \text{Runnable names} \\
  P & \in & \text{Port names} \\
  E & \in & \text{Sender-receiver element names} \\
  O & \in & \text{Client-server operation names} \\
  S & \in & \text{Inter-runnable variable names} \\
  X & \in & \text{Exclusive area names} \\
\\
  A & \in & \text{Addresses} \\
  A & ::= & \Tr{R}{I}   \sep   \Tep{E}{P}{I}   \sep \Top{O}{P}{I}   \sep   \Ts{S}{I}   \sep   \Tx{X}{I}   \\
\end{array}
\]
%
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{TODO2}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\runnable{\Tr{R}{I}}{TODOK}{T}{\Act}{N}$  \newline
Common state for runnable $R$ of component $I$, showing $N$ current instances, $T$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\rinst{\Tr{R}{I}}{C}{\Xs}{\Code}$ \newline
An instance of runnable $R$ of component $I$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $C$.
\item $\excl{\Tx{X}{I}}{B}$ \newline
Exclusive area $X$ of component $I$, with boolean occupation state $B$.
\item $\irv{\Ts{S}{I}}{V}$ \newline
Inter-runnable variable $S$ of component $I$, currently holding value $V$.
\item $\delem{\Tep{E}{P}{I}}{B}{V}$ \newline
Non-queued sender-receiver data element $E$ in port $P$ of component $I$, currently holding value $V$ with boolean update status flag $B$.
\item $\qelem{\Tep{E}{P}{I}}{\Vs}{TODO}$ \newline
Queued sender-receiver data element $E$ in port $P$ of component $I$, currently holding value sequence $\Vs$.
\item $\opres{\Top{O}{P}{I}}{\Vs}$ \newline
Client-server operation $O$ in port $P$ of component $I$, currently holding result value sequence $\Vs$.
\item $\timer{\Tr{R}{I}}{T}$ \newline  %TODO: timer has only 2 args in the semantics so {T_p} is not described
Countdown timer state for runnable $R$ of component $I$ with $T$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
\[
\begin{array}{rcll}
	V	& \in		& \text{Values} \\
	N	& \in		& \text{Natural numbers} \\
	B	& \in		& \text{Boolean values} \\
	T	& \in		& \text{Time values} \\
	C	& ::=		& \Top{O}{P}{I} \sep \Tvoid \\
	\Act	& ::=		& \Tidle \sep \Tpending \sep \serving{\Cs}{\Vs} \\
\end{array}
\]
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

The code part of a runnable instance should technically be a representation of the C or Matlab function that must accompany an AUTOSAR runnable definition once it is complete. However, since our task in this report is not to investigate the semantics of these languages in every detail, a more abstract notion of executable code will be required. To this end, we have chosen to ignore all details of internal computations, and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. Because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style to allow any data-dependencies to be expressed.
%TODO: several LaTeX commands here are used with more arguments than in the semantics below
\[
\begin{array}{lll}
	\Code	& ::=	& \Enter{X}{\Cont}			\\
			& |	& \Exit{X}{\Cont}			\\
			& |	& \IrvWrite{S}{V}{\Cont}		\\
			& |	& \IrvRead{S}{\Cont}			\\
			& |	& \Send{E}{P}{V}{\Cont}		\\
			& |	& \Receive{E}{P}{\Cont}		\\
			& |	& \Write{E}{P}{V}{\Cont}		\\
			& |	& \Read{E}{P}{\Cont}		\\
			& |	& \Invalidate{E}{P}{\Cont}		\\
			& |	& \IsUpdated{E}{P}{\Cont}		\\
			& |	& \Call{O}{P}{V}{\Cont}		\\
			& |	& \Result{O}{P}{\Cont}		\\
			& |	& \Return{V}				\\
	\\
	\Cont	& \in	& V \rightarrow \Code		\\
\end{array}
\]

Together with an operator for composing process terms in parallel, the process grammar for our AUTOSAR calculus becomes
\[
\begin{array}{rcll}
	P,Q 	& ::= &
		\runnable{A}{K}{T}{\Act}{N}	\\
		& | &
		\rinst{A}{C}{\Xs}{\Code}	\\
		& | &
		\excl{A}{B}			\\
		& | &
		\irv{A}{V}				\\
		& | &
		\delem{A}{B}{V}			\\
		& | &
		\qelem{A}{\Vs}{TODO}			\\
		& | &
		\opres{A}{\Vs}			\\
		& | &
		\timer{A}{T}		\\
		& | &
		P|Q					\\
\end{array}
\]

TODO: links from syntax and concepts introduced here to the AUTOSAR spec

@@

\item For each runnable $R$ of each component $I$, a term
\[
\begin{array}{c}
	\runnable{\Tr{R}{I}}{TODO}{0.0}{\Act}{0}
\end{array}
\]
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\Tpending$ if $R$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\Tidle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area $X$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\excl{\Tx{X}{I}}{\Tfalse}
\end{array}
\]

\item For each inter-runnable variable $S$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\irv{\Ts{S}{I}}{V}
\end{array}
\]
where $V$ is the initial value attributed to $S$ if it exists, otherwise $V$ is $\Tvoid$.

\item For each data element $E$ of each required port $P$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\qelem{\Tep{E}{P}{I}}{\epsilon}{TODO}
\end{array}
\]
if the \emph{swImplPolicy} attribute of element $E$ is \emph{queued}; otherwise, a term
\[
\begin{array}{c}
	\delem{\Tep{E}{P}{I}}{\Tfalse}{\text{nodata}}
\end{array}
\]

\item For each operation $O$ of each required port $P$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\opres{\Tep{O}{P}{I}}{\epsilon}
\end{array}
\]

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
\[
\begin{array}{c}
	\timer{\Tr{R}{I}}{0.0}
\end{array}
\]
% where $T$ is the period of the event.
%TODO: Does this mean all timed runnables will want to act _immediately_ (at the same time)? No init delay?

@@

Some further static model information will also be referenced by the transition rules. We refer to the \emph{connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
\[
	\Tconnect{\Tep{E}{$P_1$}{$I_1$}}{\Tep{E}{$P_2$}{$I_2$}} \;\;\text{for all elements $E$ of port $P_2$}
\]
if there is a connector in the model from provided sender-receiver port $P_1$ of component $I_1$ to required port $P_2$ of component $I_2$; and
\[
	\Tconnect{\Top{O}{$P_1$}{$I_1$}}{\Top{O}{$P_2$}{$I_2$}} \;\;\text{for all operations $O$ of port $P_1$}
\]
if there is a connector in the model from required client-server port $P_1$ of component $I_1$ to provided port $P_2$ of component $I_2$.

@@

\subsection{Exclusive areas}

@@
